\mytitle

### 2022-11-16 Article.tex
Revision

\def\mytitle{Whirlwind: Overload protection, fault--tolerance and self--tuning in an Internet services platform}
\def\myauthor{Peter Donald, Samar Singh and Somnath Ghosh}
\def\mykeywords{concurrency}

@@

\begin{abstract}
Performance and availability are critical importance when Internet services are integrated into emergency response management. Poor performance or service failure can result in severe economic, social or environmental cost. This paper presents Whirlwind, a software architecture that includes primitives for overload management and fault tolerance. A Whirlwind service is composed of a collection of isolated, independent, sequential processes that communicate through asynchronous message passing. If a process fails, the fault is contained within the process and a message is propagated to monitoring processes that may attempt to recover from the error. Processes are grouped with other processes that share similar resource, computation and concurrency requirements. Each group contains a scheduler and a thread pool that drives execution of processes within the group. The group may also define a message predicate that determines if a message posted to a process in the group is accepted. A rejected message typically signals overload and allows the application the chance to perform load shedding and avoid overcommitment of resources. Principles are shared between processes in different groups, enabling consistent prioritization and admission control across groups. The resource management policies are typically driven by feedback loops that monitor resource availability and system performance, and adjust tuning parameters to meet performance goals. Whirlwind evolved over a period of five fire seasons as part of emergency response software in Victoria, Australia.
\end{abstract}

@@

Internet traffic is bursty~\cite{Crovella:1996:BurstyTraffic} and popular Internet services are likely to be subject to periods of transient or sustained overload. This is intensified for Internet services used in emergency responses. Overload can be caused by unexpectedly high request rates or unexpected shifts in the characteristics of the request distribution. In IRIS the request rate increased dramatically during shift change over times or prior to payment cut--off dates. The characteristics of the requests also changed dramatically depending on whether the users were primarily interacting with payment subsystem or the resource tracking subsystem.

@@

Whirlwind is based on the actor model of concurrency. It incorporates existing work on load management, fault--tolerance, scheduling and self--tuning and synthesizes these elements into a consistent framework. Whirlwind was developed for the Java Virtual Machine, principally targeting the Java programming language. The architecture needed to take advantage of the increased number of processing units in modern hardware.

Early in the evolution of Whirlwind it was discovered that it was necessary to provide flexibility in certain aspects such as admission control policies and scheduling algorithms. Plugin points are defined where flexibility was required. To maintain acceptable performance the plugins are designed with coarse--grain interfaces with minimal virtual calls along critical paths.

@@

Load shedding is typically achieved by selectively applying the above techniques to different requests. Requests can be categorized by low level connection properties~\cite{Voigt:2001qe}, protocol specific parameters (i.e. the url, headers or session associated with a HTTP request)~\cite{Cherkasova:2002yb,Voigt:2001qe} or application specific metrics (i.e. revenue and profits~\cite{Menasec:2000ty}, or \emph{Premium} vs \emph{Basic} users). Requests can also be categorized according to the expected change in resource availability after processing request~\cite{Behren03Capriccio} or predicted resource cost to process request~\cite{Crovella99Alpha}. Other systems categorize the request based on estimated utility\cite{Zhou06RequestAware} or ability to meet performance goals (i.e. the 90th--percentile of response time~\cite{welsh03Adaptive}). Class--based differentiation is a coarse--grain classification scheme but fine--grain classification schemes are common.

@@

The construction of fault--tolerant computing systems is simplified when programs do not need to cope with arbitrary behavior or corrupted state information. So not only should a failure be contained within a process, but on detection of a failure the process should halt rather than make erroneous state visible to other processors. Other processors should be able to detect if a process has failed so that they can attempt error recovery. Error recovery should have a known good state with which to restart the process. Similar properties are deemed necessary to construct fault--tolerant hardware systems~\cite{Schneider:FailStopProcessors}.

Erlang is a software architecture designed to survive software and hardware faults~\cite{Armstrong03Thesis} and incorporates these strategies. Erlang models the world as isolated processes. When a process terminates either normally or due to failure, a message is sent to all monitoring processes. The monitoring processes can then take action to recover from the error if required.

An Erlang supervisor is a process that monitors child processes. If a child process fails, the supervisor uses a \emph{restart strategy} to restart the process and if unable to restart the process will terminate itself. Supervisors can monitor other supervisor processes and supervision is typically organized into a tree structure. The root supervisor must be reliable to ensure that the system can always resume normal operation. Erlang's supervisors were part of the architectural design that lead to 99.9999999\% (9 nines) reliability rating (or 31 ms. downtime a year!) of the AXD301 switch~\cite{armstrong:02:COP}.

@@

The looming multi--core future and the success of Erlang~\cite{armstrong96erlang} has recently placed the actor model in the spotlight. The actor model of concurrency offers an alternative to shared memory architectures. Actors are isolated and independent sequential entities that communicate by passing messages. Typical the number of actors present in a system is an order of magnitude greater than the number of threads supported.

@@

A principal is shared between one or more processes. The principal is the entity used by the scheduler and message predicates to classify and prioritize a process. Typically the processes that share a principal are in different groups and the principal enables consistent scheduling and admission control across groups.

@@

The JVM is capable of supporting an order of magnitude more processes than threads. With appropriate operating system configuration it is not unreasonable for a low end server to support 10s of 1000s of concurrent connections.

@@

Processes are isolated in and do not share memory. To communicate with a process a message is posted to the process. The ownership of the message contents is transferred to the receiving process. Message delivery is asynchronous and messages are placed in a process--local queue or inbox. Message transmission is asynchronous to ensure the the sender is isolated from the receiver and can not be blocked, or delayed by a slow receiver. It should be noted that message passing Whirlwind obeys the \emph{happened--before} ordering of events~\cite{lamport78}.

A process is not scheduled unless it has at least one message in it's inbox. A scheduled process dequeues the first message from it's inbox and passes it to the message handling code. On completion of handling the message, control is passed back to the runtime to schedule the next process. The runtime ensures that messages for a single process are handled sequentially. This design allows the process to be written without using any locks.

@@

A \emph{non--system} process that receives a \emph{non--normal} termination message will fail while normal termination messages will be silently dropped. Typically \emph{non--system} processes will have bidirectional monitoring relationship with all ``peer'' processes that share a common goal so that if one fails, all fail. For example, a HTTP connection may be represented by three or more processes; a process to manage the socket, a process to encode and decode the http protocol and a process to generate the content. If one fails then all should fail and typically the failure will also be propagated to a system process that decides whether a further error recovery response is required.

@@

Most Internet services support many concurrent requests and many of these requests perform similar computations. Identifying requests that are at a similar stage of processing and executing a ``cohort'' of them sequentially increases code and data reuse. This in turn increases performance by reducing cache misses~\cite{Larus:2002:Cohort,welsh03Adaptive}. Groups are designed to identify processes that have a high degree of code and data locality. Thus the set of processes in a group form an ideal cohort.

@@

A group may define a message predicate that offer a form of per--group admission control. Before a message is delivered to a process within the group, the message predicate may decide to reject a message. A rejected message typically signals that the group is overloaded and that there is insufficient resources to handle the message. This allows the application the chance to perform load shedding and avoid overcommitment of resources.

@@

The default scheduler uses a FIFO algorithm but there are several other schedulers included within the Whirlwind. A priority

A stride scheduler~\cite{Stride:Waldspurger:95} allocates processing proportionally to different request classes. The ability to define custom schedulers has made it possible to replicate past work such as request aware scheduling~\cite{Zhou06RequestAware} and resource aware scheduling~\cite{Behren03Capriccio}.

@@

Included with Whirlwind is a library that includes implementations of runtime components such as schedulers, message predicates, resource controllers and principles. The library also includes reusable slices of applications providing; access to file systems, access to sockets, TLS/SSL translation, HTTP connection management, schedulers, a declarative xml format for assembling and configuring applications etc. There is also a testing and benchmarking framework with a database--backed, web UI for recording and measuring performance of Whirlwind based systems.

The most interesting library included within Whirlwind is the supervisor library. The supervisor library is based loosely around the Erlang implementation of supervisors~\cite{Armstrong03Thesis}. The library includes a system process that will monitor child processes and attempt error recovery if the child processes fail. Several error recovery strategies are provided.

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{http-server.eps}
\caption{The subsystem responsible for processing HTTP connections. All of the groups contain processes of a single type except for the ``Resource Handler'' group. This group contains processes of type ``List'' for presenting a list of resources and ``Show'' for presenting detailed information about a single resource.}
\label{figure:http-server}
\end{figure*}

@@

Each request submitted to the server results in the \verb+Dispatcher+ creating a \emph{handler} process. A handler may be responsible for delivering content from the filesystem, dynamically generating content based based on values retrieved from a database or submitting data to an external system.

The handler will usually generate one or more processes while handling the request. For example, the \verb+FileHandler+ process creates an \verb+AsyncFile+ process while the \verb+ListDeployedResources+ process creates a \verb+DbConnection+ process.

Some handlers will share a group if they utilize similar resources and load shedding is likely to handle these handlers similarly. However, most handlers and other other processes in the system will have a single group per process type.

Most Whirlwind applications will only place messages predicates in a few specific locations. The HTTP server defines a message predicate on the \verb+Dispatcher+ group and on each of the handler groups.

If a message predicate attached to a handler group rejects a message then the dispatcher may attempt to send a less resource intensive request to the handler. Figure~\ref{figure:message-predicate} presents a scenario where the \verb+FileHandler+ group rejects a request for a high resolution image and a medium resolution image before accepting a request for a low resolution image.

\begin{figure}[htbp]
\centering
TODO
%\includegraphics[width=\linewidth]{http-server.eps}
\caption{TODO}
\label{figure:message-predicate}
\end{figure}

@@

Whirlwind v2.0 did not guarantee any particular event ordering. Multiple threads could be attempting to delegate event handling to a single element at any one time and the ordering is determined by the JVM or operating system. Mid fire season, the hardware platform was upgraded to 16 processors which increased the frequency of out--of--order event processing and exposed several bugs with some elements ability to handle out--of--order events.

@@

To enforce isolation properties of the messages, the byte code validator required that developer annotate the messages. The validator used the annotations combined with rules and heuristics to enforce the isolation constraints. The messages and all transitively referenced values needed to be either; immutable, a process or have ownership transferred from the source to destination processes. The byte code validator in many ways resembled a less capable, more ad--hoc version of the Kilim byte code weaver~\cite{Srinivasan:08:Kilim}.

@@

No faults were detected when Whirlwind v4.0 was deployed so the latest revision focused on performance. This resulted in the introduction of custom schedulers and principals. Custom schedulers allowed the development of application or context specific schedulers. This introduced other mechanisms to manage overload response. Principals came about as the categorization and prioritization code was being duplicated between message predicates and schedulers.

@@

\section{Related Work}

Whirlwind is an implementation of the actor model of concurrency based on the JVM that includes load management and fault management primitives. While it incorporates aspects of existing work it is most strongly related to the work on SEDA, Erlang and Kilim.

Whirlwind incorporates load protection primitives from SEDA, namely cohort scheduling and message predicates. Whirlwind also includes fault management primitives and a supervisor library inspired by similar elements in Erlang. Whirlwind, SEDA and Kilim are all JVM based while Erlang uses a specialized virtual machine.

Whirlwind eschews the shared memory model of SEDA and assumes processes are isolated and share no state. Unlike Kilim and Erlang, ensuring the isolation of processes is partially the responsibility of the developer. A developer can inadvertently or deliberately share state between processes but in such a scenario, Whirlwind assumes that the developer knows what they are doing.

At least one Whirlwind application made use of this ``feature'' and used an existing in--memory database that used fine grain locking to mediate access to the shared state access. All processes that accessed the shared state were placed in a single group. The processes were also linked so that if any failed, they all failed and were restarted.

In SEDA, events about a single subject can be handled concurrently and out--of--order while in Whirlwind messages are sequential and in a FIFO order. Both Kilim and Erlang handled messages sequentially but Erlang can choose to receive messages out of order.

None of SEDA, Erlang or Kilim allowed configuration of custom schedulers and no framework that we are aware of incorporates the principal abstraction. Only SEDA incorporates adaptive concurrency level.

Kilim and Erlang are independent threads of execution and can store state on the stack in between messages. Whirlwind and SEDA must explicitly store any continuation state in the process as they use a callback model and are have the message or event handling code invoked when a message or event arrives.

@@

There are two major changes planned for the next version of Whirlwind. The first change is to support Kilim~\cite{Srinivasan:08:Kilim} or Kilim--like processes executing on top of the Whirlwind runtime. The second is to investigate the possibility of dynamically organizing processes into groups at runtime.

@@

Secondly, Kilim processes are lightweight threads that use automatic stack management~\cite{Adya02Cooperative} to maintain continuation data between handling messages. 
Whirlwind processes currently use complex state machines and manual stack management to achieve the same effect. Kilim attempts to optimize saving and restoring stack state when a process is paused or resumed. The degree of overhead is related to the number of frames on the stack when the process is paused or resumed. Experience with existing Whirlwind applications suggest that a paused process will only have a small number of active frames. The expectation is that the small performance degradation is acceptable tradeoff for certain classes of process.

@@

Annotations could be added to the processes to provide hints to the runtime system to guide grouping. The annotations would indicate the resource and concurrency requirements of the processes. Where annotations were not sufficient for the runtime to group processes appropriately, the developer could fall back on manual group configuration.

@@

@@

Whirlwind demonstrates an architecture for building high--concurrency Internet services with overload protection and fault tolerance. Whirlwind is a synthesis of existing techniques put to the test in production environments for 5 years, in a context where there is a high cost for failure or poor performance. Whirlwind managed to withstand extreme loads as a result of Victoria's worst fire season in 2009.

The programming model introduces the principal abstraction. This centralizes the categorization and prioritization of requests and allows past work in scheduling and load shedding to be examined in a unified framework.

Whirlwind also demonstrates an approach to incorporating existing load management techniques into a framework based on the actor model of concurrency. Flexible scheduling, cohort scheduling and message predicates proved to be useful additions.

>>>>

\def\mytitle{Whirlwind: Overload protection, fault--tolerance and self--tuning in an Internet services platform}
\def\myauthor{Peter Donald, Samar Singh and Somnath Ghosh}
\def\mykeywords{concurrency, software architecture, overload, fault-tolerance, self-tuning}

@@

\begin{abstract}
Performance and availability are critical importance when Internet services are integrated into emergency response management. Poor performance or service failure can result in severe economic, social or environmental cost. This paper presents Whirlwind, a software architecture that includes primitives for overload management and fault tolerance. A Whirlwind service is composed of a collection of isolated, independent, sequential processes that communicate through asynchronous message passing. If a process fails, the fault is contained within the process and a message is propagated to monitoring processes that may attempt to recover from the error. Processes are grouped with other processes that share similar resource, computation and concurrency requirements. Each group contains a scheduler and a thread pool that drives execution of processes within the group. The group may also define a message predicate that determines if a message posted to a process in the group is accepted. A rejected message typically signals overload and allows the application the chance to perform load shedding and avoid overcommitment of resources. Principals are shared between processes in different groups, enabling consistent prioritization and admission control across groups. The resource management policies are typically driven by feedback loops that monitor resource availability and system performance, and adjust tuning parameters to meet performance goals. Whirlwind evolved over a period of five fire seasons as part of emergency response software in Victoria, Australia.
\end{abstract}

@@

Internet traffic is bursty~\cite{Crovella:1996:BurstyTraffic} and popular Internet services are likely to be subject to periods of transient or sustained overload. This is intensified for Internet services used in emergency responses. Overload can be caused by unexpectedly high request rates or unexpected shifts in the characteristics of the request distribution. In IRIS the request rate increased dramatically during shift change over times or prior to payment cut--off dates. The characteristics of the requests also changed dramatically depending on whether the users were primarily interacting with the payment subsystem or the resource tracking subsystem.

@@

Whirlwind is based on the actor model of concurrency. It incorporates existing work on load management, fault--tolerance, scheduling and self--tuning and synthesizes these elements into a consistent framework. Whirlwind was developed for the Java Virtual Machine, primarily targeting the Java programming language. The architecture needed to take advantage of the increased number of processing units in modern hardware.

Early in the evolution of Whirlwind it was discovered that it was necessary to provide flexibility in certain aspects such as admission control policies and scheduling algorithms. Plugin points are defined where flexibility is required. To maintain acceptable performance the plugins are designed with coarse--grain interfaces with minimal virtual calls along critical paths.

@@

Load shedding is typically achieved by selectively applying the above techniques to different requests. Requests can be categorized by low level connection properties~\cite{Voigt:2001qe}, protocol specific parameters (i.e. the url, headers or session associated with a HTTP request)~\cite{Cherkasova:2002yb,Voigt:2001qe} or application specific metrics (i.e. revenue and profits~\cite{Menasec:2000ty}, or \emph{Premium} vs \emph{Basic} users). Requests can also be categorized according to the expected change in resource availability after processing the request~\cite{Behren03Capriccio} or the predicted resource cost to process the request~\cite{Crovella99Alpha}. Other systems categorize the request based on the estimated utility\cite{Zhou06RequestAware} or ability to meet performance goals (i.e. the 90th--percentile of response time~\cite{welsh03Adaptive}). Class--based differentiation is a commonly used coarse--grain classification scheme but fine--grain classification schemes are also popular.

@@

The construction of a fault--tolerant computing systems is simplified when programs do not need to cope with arbitrary behavior or corrupted state information. So not only should a failure be contained within a process, but on detection of a failure the process should halt rather than make erroneous state visible to other processes. Other processes should be able to detect if a process has failed so that they can attempt error recovery. Error recovery should have a known good state with which to restart the process. Similar properties are deemed necessary to construct fault--tolerant hardware systems~\cite{Schneider:FailStopProcessors}.

Erlang is a software architecture designed to survive software and hardware faults~\cite{Armstrong03Thesis} and incorporates these strategies. Erlang models the world as a set of isolated processes. When a process terminates either normally or due to failure, a message is sent to all monitoring processes. The monitoring processes can then take action to recover from the error if required.

An Erlang supervisor is a process that monitors child processes. If a child process fails, the supervisor uses a \emph{restart strategy} to restart the process and if unable to restart the process will terminate itself. Supervisors can monitor other supervisor processes and supervision is typically organized into a tree structure. To ensure that the system can always resume normal operation the root supervisor must be reliable. Erlang's supervisors were part of the architectural design that lead to 99.9999999\% (9 nines) reliability rating (or 31 ms. downtime a year!) of the AXD301 switch~\cite{armstrong:02:COP}.

@@

The looming multi--core future and the success of Erlang~\cite{armstrong96erlang} has recently placed the actor model in the spotlight. The actor model of concurrency offers an alternative to shared memory architectures. Actors are isolated and independent sequential entities that communicate by passing messages. Typically the number of actors present in a system is an order of magnitude greater than the number of threads supported.

@@

A principal is shared between one or more processes. The principal is the entity used by the scheduler and message predicates to classify and prioritize a process. Typically the processes that share a principal are in different groups and the principal enables consistent scheduling and admission control policies across groups.

@@

The JVM is capable of supporting an order of magnitude more processes than threads. With appropriate operating system configuration it is not unreasonable for a low end server to support 100s of 1000s of concurrent actors and 10s of 1000s of concurrent connections.

@@

Processes are isolated and do not share memory. To communicate with a process a message is posted to the process. The ownership of the message contents is transferred to the receiving process. Message delivery is asynchronous and messages are placed in a process--local queue or inbox. Message transmission is asynchronous to ensure that the sender is isolated from the receiver and the sender can not be blocked, or delayed by a slow receiver. It should be noted that message passing in Whirlwind obeys the \emph{happened--before} ordering of events~\cite{lamport78}.

A process is not scheduled unless it has at least one message in it's inbox. A scheduled process dequeues the first message from it's inbox and passes it to the message handling code. On completion of handling the message, control is passed back to the runtime to schedule the next process. The runtime ensures that messages for a single process are handled sequentially. This design eliminates the need for locks.

@@

A \emph{non--system} process that receives a \emph{non--normal} termination message will fail while normal termination messages will be silently dropped. Typically \emph{non--system} processes will have bidirectional monitoring relationship with all ``peer'' processes that share a common goal so that if one fails, all fail. For example, a HTTP connection may be represented by three or more processes; a process to manage the socket, a process to encode and decode the http protocol and a process to generate the content. If one fails then all should fail and typically the failure will also be propagated to a system process that decides whether further error recovery response is required.

@@

Most Internet services support many concurrent requests and many of these requests perform similar computations. Identifying requests that are at a similar stage of processing and executing a ``cohort'' of them sequentially increases code and data locality. This in turn increases performance by reducing cache misses~\cite{Larus:2002:Cohort,welsh03Adaptive}. Groups are designed to identify processes that have a high degree of code and data locality and thus the group defines ideal ``cohort''.

@@

A group may define a message predicate that offer a form of per--group admission control. Before a message is delivered to a process within the group, the message predicate may decide to reject a message. A rejected message typically signals that the group is overloaded and that there is insufficient resources to handle the message. This gives the application the chance to perform load shedding and avoid overcommitment of resources.

@@

The default scheduler uses a FIFO algorithm but there are several other schedulers included within the Whirlwind. i.e. A stride scheduler~\cite{Stride:Waldspurger:95} allocates processing proportionally to different request classes. The ability to define custom schedulers has made it possible to replicate past work such as request aware scheduling~\cite{Zhou06RequestAware} and resource aware scheduling~\cite{Behren03Capriccio}.

@@

Included with Whirlwind is a library that includes implementations of runtime components such as schedulers, message predicates, resource controllers and principals. The library also includes reusable slices of applications providing; access to file systems, access to sockets, TLS/SSL translation, HTTP connection management, schedulers, a declarative xml format for assembling and configuring applications etc. There is also a testing and benchmarking framework with a database--backed, web UI for recording and measuring performance of Whirlwind based systems.

The most interesting library included within Whirlwind is the supervisor library. The supervisor library is loosely based on the Erlang implementation of supervisors~\cite{Armstrong03Thesis}. The library includes a system process that will monitor child processes and attempt error recovery if the child processes fail. Several error recovery strategies are provided.

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{http-server.eps}
\caption{The subsystem responsible for processing HTTP connections. All of the groups contain processes of a single type except for the ``Resource Handler'' group that includes processes of type ``List'' and ``Show'' for listing a set of resources and presenting detailed information about a single resource respectively. For the sake of simplicity, the message paths are presented as flowing in one direction but all paths are bidirectional except the handlers that communicate directly with \emph{HttpConnection} processes.}
\label{figure:http-server}
\end{figure*}

@@

Each request submitted to the server results in the \verb+Dispatcher+ creating a \emph{handler} process. A handler is responsible for delivering content from the filesystem, dynamically generating content based based on values retrieved from a database or submitting data to an external system.

The handler will usually generate one or more processes while handling the request. For example, the \verb+FileHandler+ process creates an \verb+AsyncFile+ process while the \verb+ListResources+ process creates a \verb+DbConnection+ process.

Handlers will share a group if they utilize similar resources and load shedding is likely to handle these handlers similarly. However, most handlers and other other processes in the system will have a single group per process type.

Most Whirlwind applications will only place messages predicates on a few specific groups. The HTTP server defines a message predicate on the \verb+Dispatcher+ group and on each of the handler groups. If a message predicate attached to a handler group rejects a message then the dispatcher may attempt to send a less resource intensive request to the handler, as presented in Figure~\ref{figure:message-predicate}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\linewidth]{message-predicate.eps}
\caption{The message predicate attached to the \emph{FileHandler} group rejects a message requesting high quality version of a file but accepts a message requesting a low quality version of a file that is less resource intensive.}
\label{figure:message-predicate}
\end{figure}

@@

Whirlwind v2.0 did not guarantee any particular event ordering. Multiple threads could be attempting to delegate event handling to a single element at any one time and the ordering is determined by the JVM or operating system. Mid fire season, the hardware platform was upgraded to 16 processors which increased the frequency of out--of--order event processing and exposed several bugs with some elements ability to handle reordered events.

@@

To enforce isolation properties of the messages, the byte code validator required that developer annotate the messages. The validator used the annotations combined with rules and heuristics to enforce the isolation constraints. The messages and all transitively referenced values needed to be either; immutable, a process or have ownership transferred from the source to destination processes. The byte code validator in many ways resembled a less capable, more ad--hoc version of the Kilim byte code weaver~\cite{Srinivasan:08:Kilim} that just performed validation.

@@

No faults were detected when Whirlwind v4.0 was deployed so the latest revision focused on performance. This resulted in the introduction of custom schedulers and principals. Custom schedulers allowed the development of application or context specific schedulers. This introduced another mechanism to manage overload response. Principals came about as the categorization and prioritization code was being duplicated between message predicates and schedulers.

@@

\subsection{Influences}

Three major influences on Whirlwind are SEDA, Erlang and Kilim. Whirlwind incorporates; the load protection primitives from SEDA, fault management primitives and a supervisor library inspired by Erlang, and is an implementation of the actor model of concurrency for the Java programming language like Kilim. Whirlwind applications have a distinctly different flavor from any of these architectures due to different set of trade--offs.

Whirlwind eschews the shared memory model of SEDA and assumes processes are isolated and share no state. Unlike in Kilim and Erlang, ensuring the isolation of Whirlwind processes is partially the responsibility of the developer. A developer can inadvertently or deliberately share state between processes but in such a scenario, Whirlwind assumes that the developer knows what they are doing.

In SEDA, events about a single subject can be handled concurrently and can be reordered while in Whirlwind messages are processed sequentially in a FIFO order. Kilim and Erlang handle messages sequentially but Erlang need not receive messages from the inbox in a FIFO order.

Cohort scheduling, adaptive concurrency levels and message predicates are present in Whirlwind and SEDA but are not built in to Erlang or Kilim. None of these architectures make it easy to supply custom schedulers and no framework that we are aware of incorporates the principal abstraction.

Kilim and Erlang actors are independent threads of execution and can store state on the stack in between messages. Whirlwind and SEDA use a callback model and must explicitly store any continuation state in the process.

Lastly Erlang uses a specialized virtual machine to execute programs. The other architectures are all JVM based.

@@

There are three major changes planned for the next version of Whirlwind. The first change is to support Kilim~\cite{Srinivasan:08:Kilim} or Kilim--like processes executing on top of the Whirlwind runtime. The second is to investigate the possibility of dynamically organizing processes into groups at runtime. Lastly, Whirlwind should support alternative concurrency models for processes.

@@

Secondly, Kilim processes are lightweight threads that use automatic stack management~\cite{Adya02Cooperative} to maintain continuation data between handling messages. 
Whirlwind processes currently use complex state machines and manual stack management to achieve the same effect. Kilim attempts to optimize saving and restoring stack state when a process is paused or resumed. The degree of overhead is related to the number of frames on the stack when the process is paused or resumed. Experience with existing Whirlwind applications suggest that a paused process will only have a small number of active frames. The expectation is that the small performance degradation is an acceptable tradeoff for certain classes of process.

@@

Annotations could be added to the processes to provide hints to the runtime system to guide grouping. The annotations would indicate the resource and concurrency requirements of the processes. Where annotations are not sufficient for the runtime to group processes appropriately, the developer can fall back on manual group configuration.

@@

\subsection{Alternative Concurrency Models}

At least one production application breaks the isolation assumptions of Whirlwind and shares state between processes. The shared state resulted from using an existing, high performance in--memory database library. The database library used fine--grain locking to mediate access to the shared state and maintains a high level of concurrency and throughput for the common case of no contention. Wrapping the database library in a single process resulted in a significant decrease in performance and was considered non--viable. The decision was made to share the state across several processes but move all the processes to a separate group and add error propagation paths between all processes.

A process that acquires a lock can influence the behavior of another process but that process is guaranteed to be in the same group. The groups scheduler and adaptive concurrency level controller were customized to take this into account when making resource allocation decisions.

Adding error propagation paths between all processes ensures that if any process fails, they all fail. This isolates the fault to the group but does mean that a single failure can result in the error propagating to a large number of dependent processes.

These steps ensured that the group is isolated from the rest of the system even if the processes within the group are not isolated from each other.

A pragmatic addition to Whirlwind would be to enable this design at the framework level. The framework would use one execution strategy when dealing with isolated message--driven processes and another when processes can also communicate via locks and shared memory. It is conceivable that this could be extended to processes that use shared transactional memory.

@@

Whirlwind demonstrates that existing load management techniques can be incorporated into a framework based on the actor model of concurrency. Flexible scheduling policies, cohort scheduling and message predicates proved to be useful additions to the actor runtime.

The programming model introduces the principal abstraction. The principal centralizes the categorization and prioritization of requests and allows past work in scheduling and load shedding to be examined in a unified framework.

Whirlwind is a synthesis of existing techniques put to the test in a production environment for five years, in a context where there is a high cost for failure or poor performance. Several Whirlwind--based emergency response systems experienced extreme loads as a result of Victoria's worst fire season in 2009. The overload was effectively managed due in part to the infrastructure provided by Whirlwind.
